\bf Model Visualization Techniques for a Social Network Model

### 2017-08-18 paper.Rnw
through sec 3.2

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, the objective function is written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

@@

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above. \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, and are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

@@

When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes. In \citet{modelsSnijders}, the author recommends that the $U_i(x)$ be random draws from a type 1 extreme value distribution on top of the objective function. % XX What is the thought behind that? } ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p. 368]{snijders01}. 
This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of

\begin{equation} \label{eq:dist1}
f(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + e^{-\frac{u-\mu}{\sigma}}\right)\right\}.
\end{equation}

Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because its \st{properties lead to a probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written only} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, where the tie $x_{ij}$ has changed to $1-x_{ij}$, as $x(i \leadsto j)$. Then, the probility that the tie $x_{ij}$ changes is

@@

\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

@@

\st{XXX TO DO XXX}

>>>>

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, its objective function, which it tries to maximize, is written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

@@

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. \st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks. The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above. The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change. They hope to answer questions such as, ``How does the existing network structure influence change in the network?" \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest. The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

@@

When node $i$ is given the chance to change a tie, %\hh{`we assume that they wish' sounds strange - don't assume wishes, state assumptions}
it attempts to maximize the value of its objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$, to account for unknown attraction between nodes. In \citet{modelsSnijders}, it is recommended that the $U_i(x)$ be random draws from a type 1 extreme value distribution. The additional random element is included to account for any random, unexplainable change in the network ties. % XX What is the thought behind that? } ``the type 1 extreme value distribution (or Gumbel distribution) with mean 0 and scale parameter 1" \citep[p. 368]{snijders01}. 
This distribution, which is also known as the log-Weibull distribution, has probability distribution function, using $\mu$ for the mean parameter and $\sigma$ for the scale parameter, of

\begin{equation} \label{eq:dist1}
g(u|\mu, \sigma) = \frac{1}{\sigma}\exp\left\{-\left(\frac{u-\mu}{\sigma} + \exp^{-\frac{u-\mu}{\sigma}}\right)\right\}.
\end{equation}

<<logweibull, fig.align='center', out.width = '.5\\linewidth', fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter \$\\mu = 0\$ and scale parameter \$\\sigma = 0\$.'>>=
library(reliaR)
x <- seq(-3,10,.01)
y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
qplot(x = x, y= y, geom = "line") + 
  annotate('text', x = 6, y = .25, 
        label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
  labs(x = "u", y = "g(u) (density)")
@

The probability density function for the type 1 extreme value distribution is shown in Figure~\ref{fig:logweibull}. Using the distribution given in Equation~\ref{eq:dist1} with mean $\mu = 0$ and scale $\sigma = 1$ as \citet{modelsSnijders} suggests \hh{XXX I'm not following - which distribution do you mean by `this' distribution? The density above is the type 1 extreme value density. I got lost at the  ``objective function $f_i(\boldsymbol{\beta}, x)$ plus a random element, $U_i(x)$"} \st{XXX I hope it is clearer now above}
is convenient because it \st{leads to a simple probability formula for} the probability that actor $i$ chooses to change its tie to actor $j$ \st{that can be written \textit{only}} in terms of the objective function. Let $p_{ij}(\boldsymbol{\beta}, x)$ be the probability that actor $i$ chooses to change its tie to actor $j$. Next, we write the network $x$ in its potential future state, $x(i \leadsto j)$, where the tie $x_{ij}$ has changed to $1-x_{ij}$. Then, the probility that the tie $x_{ij}$ changes is

@@

When $i = j$ in $p_{ij}$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. When the value of the objective function is high at the current state, the probability of not making a change in a microstep is also high. In the CTMC, when actor $i$ may make a change, it chooses which tie $x_{i1, \dots, x_{in}}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x)$.
\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

@@

In the continuous time Markov chain literature, the intensity function... see \citet{ctmc}
